Ten Simple Rules for Teaching Programming

### 2017-07-21 paper.tex
Added a reference to rule 9.
One more pass through the paper; various minor changes.

Research from educational psychology suggests that teaching and learning are subject-specific
activities~\cite{mayer-subject}: learning programming has a different set of challenges and techniques
than learning physics or learning to read and write.  Computing is a younger discipline than
mathematics, physics, or biology, and while there have been correspondingly fewer studies of how best to teach it,
there is a growing body of evidence about what works and doesn't.
This paper presents ten simple rules that should be the foundation of any teaching of programming, whether
formal or informal.

@@

The most powerful evidence for this comes from Patitsas et al.~\cite{patitsas-bimodal}.
They examined grade distributions in introductory computing courses at a large university,
and found that only 5.8\% were actually multi-modal.
More damningly,
they found that computer science faculty were more likely to see distributions as bimodal
if they thought those grades came from a programming class
than if they believed the grades came from some other kind of class,
and that those faculty were even more likely to see the distributions as bimodal
if they believed some students are innately predisposed to do well in computer science.

Beliefs such as this are known to have powerful effects on education outcomes
\cite{alvidrez-weinstein,brophy,jussim-eccles}.
If instructors believe that ``some kids get it and some kids don't'',
they will (consciously or unconsciously) invest less in those whom they put in the second category.
When combined with cultural stereotypes about who is and isn't a "natural programmer",
the downward spiral of under-achievement that results from differential attention
may be partly responsible for the gender imbalance in computing.

@@

\item The instructor then gives learners a multiple choice question
  that probes for misconceptions rather than simple factual recall.
  A programming example is given in Figure~\ref{fig:mcq}
  which relates to integer comparison and loops.
  There is no point asking a trivial question that all students will get right.
  The ideal questions are those where students are most likely to vote equally for each of the options.

@@

  \begin{itemize}
  \item If they all have the right answer, move on.
  \item If they all have the same wrong answer,
    the instructor addresses that specific misconception.
  \item If they have a mix of right and wrong answers,
    they are given several minutes to discuss those answers with one another
    in small groups (typically 2-4 students)
    and then reconvene and vote again.
  \end{itemize}

@@

\item It enables instructors to be more responsive to ``what if?''
  questions. Where a slide deck is like a highway, live coding
  allows instructors to go off road and follow their learners'
  interests.

\item It facilitates unintended knowledge transfer: students learn
  more than the instructor consciously intends to teach by watching
  \textit{how} instructors do things.  The extra knowledge may be
  high-level (e.g., whether a program is written top-down or
  bottom-up) or fairly low-level (e.g., using keyboard shortcuts).

\item It slows the instructor down: if the instructor has to type in
  the program as they go along, they can only go twice as fast as
  their learners, rather than ten-fold faster as they could with
  slides.

@@

\item Instructors can spend too much time typing in boilerplate code
  that is needed by the lesson, but not directly relevant to it (such
  as library import statements), can distract learners from the
  intended thrust of a lesson.  As Willingham~\cite{willingham-book} says,
  ``Memory is the residue of thought''; if the instructor spends their
  time typing boilerplate, that may be all that learners take away.
  This can be avoided by starting with a partial skeleton that
  includes the boilerplate, or having it on hand to copy and paste
  when needed.  (Of the two, we prefer the former, since learners
  may not be able to keep up with copying and pasting.)

@@

When instructors are using live coding, they usually run the program
several times during its development to show what it does.  Research
from peer instruction in physics education shows that learners who
observe a demonstration \textit{do not} learn better than those who
did not see the demonstration~\cite{crouch-demo}, and in fact many
learners misremember the outcome of demonstrations
afterwards~\cite{miller-demo}.  In other words, demonstrations can
actually be useless or actively harmful.

The key to making demonstrations more effective is
to make learners predict the outcome of the demonstration before performing it.
Crucially, their prediction should be in some way recorded or public,
e.g. by a show of hands,
by holding up a cue card (A/B/C/D),
or by talking to their neighbour.
We speculate that the sting of being publicly wrong leads learners to pay more attention
and to reflect on what they are learning;
regardless of whether this hypothesis is true,
instructors should be careful not to punish or criticise students who predicted wrongly,
but rather to use those incorrect predictions as a spur to further exploration.

@@

Instructors usually provide many similar programming examples for learners to practice on.
But since learners are novices,
they may not see the similarity between examples:
finding the highest rainfall from a list of numbers and
finding the first surname alphabetically from a list of names
may seem like quite different problems to them,
even though more advanced programmers would recognise them as isomorphic.

Morrison et al.~\cite{morrison-subgoals,morrison-parsons,margulieux} have shown that students perform better
when worked examples are broken down into steps (or subgoals)
which are given names (or labels) -- an example is given in Figure~\ref{fig:subgoal}.
Subgoal labels provide a structure which students can apply to future tasks that they attempt themselves,
and allow them to communicate with their peers and instructors more efficiently.

@@

One caution is that context can inadvertently exclude some people while drawing others in.
For example,
many educators use computer games as a motivating example for programming classes,
but some learners may associate them with violence and racial or gender stereotypes,
or simply find them unenjoyable.
Whatever examples are chosen,
the goal must be to move learners as quickly as possible
from ``hard and boring'' to ``easy and exciting'' \cite{repenning}.

@@

This principle is tautological, but it is easily forgotten.
Novices program differently than experts,
and need different approaches or tools.
If you ask a professional programmer to iterate over a list of integers and produce the average,
they can write the code within seconds,
using stored knowledge of the exact pattern required.
A novice will approach this problem totally differently:
they need to remember the syntax for the different parts,
they need to know how to iterate over a list,
how to use an accumulator variable,
and so on.

@@

Our final rule for teaching programming is that you don't have to program to do it.
Between syntax, semantics, algorithms, and design,
examples that seem small to instructors can easily overwhelm novices.
Breaking the problem down into smaller single-concept pieces
can reduce the cognitive load to something manageable.

For example,
a growing number of educators are including Parsons Problems
in their pedagogic repertoire~\cite{parsons,morrison-parsons}.
Rather than writing programs from scratch,
learners are given the lines of code they need to solve a problem,
but in jumbled order.
Re-ordering them to solve the problem correctly
allows them to concentrate on mastering control flow
without having to devote mental energy to recalling syntax
or the specifics of library functions.

@@

Like anything involving human subjects,
studies of computing education must necessarily be hedged with qualifiers.
However,
we do know a great deal,
and are learning more each year.
Venues like SIGCSE (\url{http://sigcse.org/}), ITiCSE (\url{http://iticse.acm.org/}) and ICER (\url{https://icer.hosting.acm.org})
present a growing number of rigorous, insightful studies
with immediate practical application.
Future work may overturn or qualify some of our ten simple rules,
but they form a solid basis for any educational effort.

We offer one final observation: do not forget the human element.
Programmers have a reputation for pouring scorn on certain programming languages (e.g., PHP),
or for gatekeeping (e.g., stating that you can't learn programming if you didn't start young).
If you are teaching someone to program,
the last thing you want to do is make them feel like they can't succeed
or that any existing skill they have (no matter when or how acquired) is worthless.
Make your learners feel that they can be a programmer,
and they just might become one.

>>>>

Research from educational psychology suggests that teaching and learning are subject-specific
activities~\cite{mayer-subject}: learning programming has a different set of challenges and techniques
than learning physics or learning to read and write.  Computing is a younger discipline than
mathematics, physics, or biology, and while there have been correspondingly fewer studies of how best to teach it,
there is a growing body of evidence about what works and what doesn't.
This paper presents ten simple rules that should be the foundation of any teaching of programming, whether
formal or informal.

@@

The most powerful evidence for this comes from Patitsas et al.~\cite{patitsas-bimodal}.
They examined grade distributions in introductory computing courses at a large university,
and found that only 5.8\% were actually multi-modal.
More damningly,
they found that computer science faculty were more likely to see distributions as bimodal
if they thought those grades came from a programming class
than if they believed the grades came from some other kind of class,
and that those faculty were even more likely to see the distributions as bimodal
if they believed that some students are innately predisposed to do well in computer science.

Beliefs such as this are known to have powerful effects on education outcomes
\cite{alvidrez-weinstein,brophy,jussim-eccles}.
If instructors believe that ``some kids get it and some kids don't'',
they will (consciously or unconsciously) invest less in those whom they put in the second category.
When combined with cultural stereotypes about who is and isn't a ``natural programmer'',
the downward spiral of under-achievement that results from differential attention
may be partly responsible for the gender imbalance in computing.

@@

\item The instructor then gives learners a multiple choice question
  that probes for misconceptions rather than simple factual recall.
  A programming example is given in Figure~\ref{fig:mcq}
  which relates to integer comparison and loops.  The multiple choice question must be well designed.
  There is no point asking a trivial question that all students will get right or one with meaningless wrong answers which no student will pick.
  The ideal questions are those where 40--60\% of students are likely to get the right answer first time~\cite[p23]{nrc-reaching}, and where every wrong answer corresponds to a misconception that will cause it to be picked by at least some students.

@@

  \begin{itemize}
  \item If they all have the right answer, the instructor can move on.
  \item If they all have the same wrong answer,
    the instructor addresses that specific misconception.
  \item If they have a mix of right and wrong answers,
    they are given several minutes to discuss those answers with one another
    in small groups (typically 2-4 students)
    and then reconvene and vote again.
  \end{itemize}

@@

\item It enables instructors to be more responsive to ``what if?''
  questions. Where a slide deck is like a highway, live coding
  allows instructors to go off road and follow their learners'
  interests or answer unanticipated questions.

\item It facilitates unintended knowledge transfer: students learn
  more than the instructor consciously intends to teach by watching
  \textit{how} instructors do things.  The extra knowledge may be
  high-level (e.g., whether a program is written top-down or
  bottom-up) or fairly low-level (e.g., learning useful editor shortcuts).

\item It slows the instructor down: if the instructor has to type in
  the program as they go along, they can only go twice as fast as
  their learners, rather than ten-fold faster as they could with
  slides -- which risks leaving everyone behind.

@@

\item Instructors can spend too much time typing in boilerplate code
  that is needed by the lesson, but not directly relevant to it (such
  as library import statements). Not only does this slow things down, it can distract learners from the
  intended thrust of a lesson.  As Willingham~\cite{willingham-book} says,
  ``Memory is the residue of thought''; if the instructor spends their
  time typing boilerplate, that may be all that learners take away.
  This can be avoided by starting with a partial skeleton that
  includes the boilerplate, or having it on hand to copy and paste
  when needed.  (Of the two, we prefer the former, since learners
  may not be able to keep up with copying and pasting.)

@@

When instructors are using live coding, they usually run the program
several times during its development to show what it does.  Surprising research
from peer instruction in physics education shows that learners who
observe a demonstration \textit{do not learn better} than those who
did not see the demonstration~\cite{crouch-demo}, and in fact many
learners misremember the outcome of demonstrations
afterwards~\cite{miller-demo}.  In other words, demonstrations can
actually be useless or actively harmful.

The key to making demonstrations more effective is
to make learners predict the outcome of the demonstration before performing it.
Crucially, their prediction should be in some way recorded or public,
e.g. by a show of hands,
by holding up a cue card (A/B/C/D),
or by talking to their neighbor.
We speculate that the sting of being publicly wrong leads learners to pay more attention
and to reflect on what they are learning;
regardless of whether this hypothesis is true,
instructors should be careful not to punish or criticise students who predicted wrongly,
but rather to use those incorrect predictions as a spur to further exploration and explanation.

@@

Instructors usually provide many similar programming examples for learners to practice on.
But since learners are novices,
they may not see the similarity between examples:
finding the highest rainfall from a list of numbers and
finding the first surname alphabetically from a list of names
may seem like quite different problems to learners,
even though more advanced programmers would recognise them as isomorphic.

Morrison et al.~\cite{morrison-subgoals,morrison-parsons,margulieux} have shown that students perform better
when worked examples are broken down into steps (or subgoals)
which are given names (or labels) -- an example is given in Figure~\ref{fig:subgoal}.
Subgoal labels provide a structure which allow learners to see the similarities between coding problems and to communicate with their peers and instructors more efficiently.  Learners can then apply the labels to future tasks that they attempt themselves.

@@

One caution about choosing context is that context can inadvertently exclude some people while drawing others in.
For example,
many educators use computer games as a motivating example for programming classes,
but some learners may associate them with violence and racial or gender stereotypes,
or simply find them unenjoyable.
Whatever examples are chosen,
the goal must be to move learners as quickly as possible
from ``hard and boring'' to ``easy and exciting'' \cite{repenning}.

To help students accomplish a visible and satisfting result
quickly, instructors can provide some pre-written software libraries or source code that start students
closer to the end goal.  The idea that students must start from scratch and write all the code they need
themselves is the relic of a bygone era of home microcomputers (and it was not true even then).
Pick the task that you actually want to the students to engage in, and provide everything else pre-made.

@@

This principle is tautological, but it is easily forgotten.
Novices program differently than experts~\cite{parnin-expertise},
and need different approaches or tools.
If you ask a professional programmer to iterate over a list of integers and produce the average,
they can write the code within seconds,
using stored knowledge of the exact pattern required.
A novice will approach this problem totally differently:
they need to remember the syntax for the different parts,
they need to know how to iterate over a list,
how to use an accumulator variable,
and so on.

@@

Our final rule for teaching programming is that you don't have to program to do it.
Faced with the challenges of learning syntax, semantics, algorithms, and design,
examples that seem small to instructors can still easily overwhelm novices.
Breaking the problem down into smaller single-concept pieces
can reduce the cognitive load to something manageable.

For example,
a growing number of educators are including Parsons Problems
in their pedagogic repertoire~\cite{parsons,morrison-parsons}.
Rather than writing programs from scratch,
learners are given the lines of code they need to solve a problem,
but in jumbled order.
Re-ordering them to solve the problem correctly
allows them to concentrate on mastering control flow
without having to devote mental energy to recalling syntax
or the specifics of library functions.  They are also liked by learners; Ericson et al.~\cite{ericson-ebook} found
that learners were more likely to attempt Parsons Problems than nearby multiple choice questions in an ebook.

@@

The ten rules presented here are backed up by scientific research.
Like any research involving human subjects,
studies of computing education must necessarily be hedged with qualifiers.
However,
we do know a great deal,
and are learning more each year.
Venues like SIGCSE (\url{http://sigcse.org/}), ITiCSE (\url{http://iticse.acm.org/}) and ICER (\url{https://icer.hosting.acm.org})
present a growing number of rigorous, insightful studies
with immediate practical application.
Future work may overturn or qualify some of our ten simple rules,
but they form a solid basis for any educational effort to the best of our current knowledge.

We offer one final observation: do not forget the human element.
Programmers have a reputation for pouring scorn on certain programming languages (e.g., PHP),
or for gatekeeping (e.g., stating that you cannot learn programming if you did not start young).
If you are teaching someone to program,
the last thing you want to do is make them feel like they can't succeed
or that any existing skill they have (no matter when or how acquired) is worthless.
Make your learners feel that they can be a programmer,
and they just might become one.

\section*{Acknowledgments}

We are grateful to Barbara Ericson for her comments on a draft of this article.
